---
title: "NCAA Budgets"
author: "Sean Li"
date: "2023-12-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r packages}
library(tidyverse)
library(kableExtra)
library(tidytuesdayR)
```

```{r load-data}
sports <- read.csv("sports.csv")
```



##Introduction

### Background

College sports in the United States bring in over 15.8 billion dollars in revenue a year according to the NCAA in 2019. The lions share of those revenue come from football - with each school averaging 31.9 million dollars per school annually. These dollars come mainly from the huge TV streaming rights such as ESPN. The dominance of college football in revenue generation can be shown with a quick comparison. Duke University, one of the top programs in the 2nd most revenue generating sport of men's basketball, brings in approximately 37 million while top football programs such as Alabama or Georgia reel in over 3 times that, at about 110 million dollars. 

However, all these revenue dollars doesn't equate to a profitability. In 2019, only around 20% schools in the high grossing Football Bowl Subdivision (FBS) report a net positive operating margin. Main expense items include 35% to administrative and coaches, 18% on athletic facilities, and 19% to student athletes themselves in the form of scholarships. Although the expenditure has also been steadily increasing over the last few years, athletic departments are not too concerned. 

Unlike professional sports, athletic departments do not have bottom line-profits and stockholders that pressure them; instead the stakeholders such as alumni, students, and boosters want victories. Aside from just being good publicity for the athletics at the school, it increases the recruiting of top athletes (who generate more revenue and victories), contribute to image and # of applications, and also can positively impact alumni donations. This makes our analysis ever more interesting, as it deviates from the traditional sports business incentive alignments slightly. While this fact might be interesting, it does not deter the aspirations for lucrative sports programs across the United States college campuses. We still have plenty reason to specifically analyze the spending and revenues of college sports.

The landscape of NCAA sports has undergone dynamic transformations in recent years, with a central focus on the financial implications of collegiate athletics. As colleges navigate the intricate web of NCAA regulations, the impact of recent conference realignments further amplifies the complexities surrounding sports budgets. Conference shifts not only reshape competitive dynamics but also pose significant financial challenges and opportunities, with schools strategically positioning themselves to maximize revenue potential. 

### Objectives and Goals

The future of college football is changing with conferences merging and the playoff structure. How will this potentially affect spending and projected revenue? What patterns currently exist from the ACC or SEC in revenue that could allow us to identify key reasons for moving conferences? Are there specific schools that might move due to football profitability concerns? 

This study delves into the intricate relationship between NCAA sports budgets, particularly football spending, analyzing the repercussions of recent conference realignments on the fiscal trajectories of collegiate athletics.

Understanding these dynamics is key for institutions seeking to optimize their financial strategies and adapt to the evolving paradigm of collegiate sports.

### EDA 

```{r EDA}

```


## Methodology


## Results

## Discussion





## Appendix 

### Works Cited

https://athleticdirectoru.com/articles/revenue-per-quality-of-college-football-recruit/#:~:text=The%20average%20annual%20total%20revenue,over%20%2435%20million%20in%20revenue.

https://ncaaorg.s3.amazonaws.com/research/Finances/2020RES_D1-RevExp_Report.pdf 

https://www.pbs.org/newshour/economy/analysis-who-is-winning-in-the-high-revenue-world-of-college-sports 